Concurrent Programming with CXL

### 2020-05-21 book/paper.tex
More on reader/writer lock
Trying to tell why busy wait is bad...

What we want is a special kind of lock that allows either one writer
or one or more readers to be in the critical section.  This is called
a \emph{reader/writer lock}.
We will explore various ways of implementing reader/writer locks in
this chapter and future ones.

@@

A reader process that wants to enter \texttt{@rcs} first waits
until there are no writers: \texttt{nwriters == 0}.
If so, it increments the number of readers.  Similarly, a writer
process waits until there are no readers \emph{or} writers.
If so, it increments the number of writers.
The important invariants in this code are:
\begin{itemize}
\item $n$ readers at \texttt{@rcs} $\Rightarrow \mathtt{nreaders} = n$,
\item $n$ writers at \texttt{@rcs} $\Rightarrow \mathtt{nwriters} = n$,
\item $(\mathtt{nreaders} \ge 0 \land \mathtt{nwriters} = 0) \lor
    (\mathtt{nreaders} = 0 \land 0 \le \mathtt{nwriters} \le 1)$.
\end{itemize}
It is easy to see that the invariants hold and imply the specification.

>>>>

What we want is a special kind of lock that allows either one writer
or one or more readers to be in the critical section.  This is called
a \emph{reader/writer lock}~\cite{courtois71}.
We will explore various ways of implementing reader/writer locks in
this chapter and future ones.

@@

A reader process that wants to enter \texttt{@rcs} first waits
until there are no writers: \texttt{nwriters == 0}.
If so, it increments the number of readers.  Similarly, a writer
process waits until there are no readers \emph{or} writers.
If so, it increments the number of writers.
The important invariants in this code are:
\begin{itemize}
\item $n$ readers at \texttt{@rcs} $\Rightarrow \mathtt{nreaders} = n$,
\item $n$ writers at \texttt{@rcs} $\Rightarrow \mathtt{nwriters} = n$,
\item $(\mathtt{nreaders} \ge 0 \land \mathtt{nwriters} = 0) \lor
    (\mathtt{nreaders} = 0 \land 0 \le \mathtt{nwriters} \le 1)$.
\end{itemize}
It is easy to see that the invariants hold and imply the specification.
The solution also supports progress: if no process is in the critical
section then any process can enter.

While correct, it is not considered a good solution.
The solution is an example of what is called \emph{busy waiting}:
processes spin in a loop until some desirable condition is met.
The astute reader might wonder if obtaining an ordinary lock itself is an
example of busy waiting.  After all, the CXL \texttt{synch} implementation
of \texttt{lock()} spins in a loop until the lock is available
(see Figure~\ref{fig:locks}).

In most operating systems and programming language runtimes, however,
when a process acquires a lock, the process is placed on a scheduling
queue and stops using CPU cycles until the lock becomes available.
Moreover, if there are multiple processes waiting for a lock, there
is a scheduler that decides which of the processes gets the lock.
Busy waiting disables all this.

But even when multiple processes each run on their own core, the solution
is inefficient.  After all, when the lock becomes available, the process
has to check other variables to see if it can continue.  In our case,
the process may have to read \texttt{nreaders} and/or \texttt{nwriters}
to decide if it can go on or has to try again, involving releasing
the lock and re-acquiring it.

So it is considered ok to have processes busy wait for a lock to become
available, but not to busy wait for application-specific conditions.
